---
title: "Introduction to EGRET Confidence Intervals"
author: "Robert M. Hirsch, Laura A. De Cicco"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Introduction to EGRET Confidence Intervals}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
library(rmarkdown)
options(continue=" ")
options(width=60)
library(knitr)
library(EGRET)
library(EGRETci)

```

# Introduction

This package **EGRETci** implements a set of approaches to the analysis of uncertainty associated with WRTDS trend analysis as implemented in the **EGRET** package.  The various functions included here are all discussed in the _A bootstrap method for estimating uncertainty of water quality trends_ by Hirsch, Archfield, and De Cicco which has been accepted for publication in the journal Environmental Modelling and Software.  As soon as it is published on line (late summer to early fall, 2015) the paper will be available at the EGRET web page at https://github.com/USGS-R/EGRET/wiki.  The  **EGRETci** package is designed for interactive or batch use.  Instructions for both are included here in the vignette.  The package allows users to set the Period of Analysis, and various smoothing parameters for WRTDS.  Users should consult the **EGRET** vignette for more background on those issues. 

  The **EGRETci** package is designed to carry out three specific types of tasks.
    
    
1)  Evaluate a water quality trend over a specific span of years and produce a variety of tabular results.  This is done with a short workflow involving the functions: `trendSetUp` and `wBT`.  The results come in three forms: 1) console output, which shows the bootstrap replicate process as it is underway and the results when it has finished, 2) a text file that shows the results of the bootstrap analysis (a subset of what is included in the console output), and 3) a set of outputs in a named list called eBoot.  The contents of eBoot are described below.  

2)  Plot histograms of values for the trend magnitudes, expressed in percent change over the specified period, for flow-normalized concentration and flow-normalized flux.  This is done with the function `plotHistogramTrend`.  It depends on outputs contained in eBoot.  Note that there are a number of custom outputs similar to these histograms that can be developed from the contents of eBoot (for example, what is the likelihood that the flow normalized flux decreased by more than 2000 kg/year over the trend period).  Such analyses would require a small amount of script writing by the user.
 


3)  Plot confidence bands around the computed trends in flow-normalized concentration and flow-normalized flux.  This is done using a function called `ciCalculations` and then, using the output from that function running two functions that produce the confidence band graphics for concentration and flux respectively (`plotConcHistBoot`, and `plotFluxHistBoot`).

## Required Data Frames

The following table provides definitions of the four data frames that constitute the named list **eBoot**

|Data Frame |Column | Definition |
|:---------|:----|:----|
|bootOut|  rejectC  |Reject Ho, (no trend in concentration), TRUE or FALSE |  
|| pValC | two-sided p-value for no trend in concentration | 
|| estC | standard WRTDS estimate of change from starting year to ending year in mg/L | 
|| lowC90 | Lower confidence limit (90%) on concentration trend | 
|| upC90 | Upper confidence limit (90%) on concentration trend | 
|| lowC50 | Lower confidence limit (50%) on concentration trend | 
|| upC50 | Upper confidence limit (50%) on concentration trend |
|| lowC95 | Lower confidence limit (95%) on concentration trend | 
|| upC95 | Upper confidence limit (95%) on concentration trend | 
|| likeCUp | Likelihood that trend in concentration is upwards | 
|| likeCDown | Likelihood that trend in concentration is downwards |
|| rejectF | Reject Ho, (no trend in flux), TRUE or FALSE | 
|| pValF | two-sided p-value for no trend in flux |
|| estF | estimate of change from starting year to ending year, in 10^6^ kg/yr |
|| lowF90 | Lower confidence limit (90%) on flux trend |
|| upF90 | Upper confidence limit (90%) on flux trend |
|| lowF50 | Lower confidence limit (50%) on flux trend |
|| upF50 | Upper confidence limit (50%) on flux trend |
|| lowF95 | Lower confidence limit (95%) on flux trend |
|| upF95 | Upper confidence limit (95%) on flux trend |
|| likeFUp | Likelihood that trend in flux is upwards |
|| likeFDown | Likelihood that trend in flux is downwards |
|| baseConc | Estimated mean flow-normalized concentration for first year, in mg/L |
|| baseFlux | Estimated mean flow-normalized flux for start year, in 10^6^ kg/yr |
|| iBoot | The actual number of bootstrap replicates used |
|wordsOut | | a vector of four character variables (self explanatory) |
|xConc | | a vector of length iBoot, of the change in flow normalized concentration computed by each bootstrap replicate (mg/L) |
|xFlux | | a vector of length iBoot, of the change in flow normalized flux computed by each bootstrap replicate (10^6^ kg/yr) |
|pConc | | a vector of length iBoot, of the change in flow normalized concentration computed from each bootstrap replicate expressed as % change |
|pFlux | | a vector of length iBoot, of the change in flow normalized concentration computed from each bootstrap replicate expressed as % change |



There is also a data frame called caseSetUp (created with the `trendSetUp` function), which contains a number of important parameters that define the way that the test was implemented.  They are presented here.

|Column |Definition |
|:----|:----|
|year1 | the water year that is the start of the trend period (an integer) | 
|yearData1 | the water year that is the start of the data set (an integer) | 
|year2 | the water year that is the end of the trend period (an integer) |
|yearData2 | the water year that is the end of the data set (an integer) |
|numSamples | number of samples in eList$Sample |
|nBoot | maximum number of replicates (called Mmax in paper) |
|bootBreak | minimum number of replicates (called Mmin in paper) |
|blockLength | length of blocks for bootstrap (called B in the paper) |
|confStop | 1 - alphap, the width of the confidence interval used in adaptive stopping rule (default alphap=0.3 so confStop=0.7) |


# WRTDS Bootstrap Test

The WBT (WRTDS Bootstrap Test) can be run interactively or in batch.  

Regardless of which approach is used, if the user wishes to use any non-default values of the parameters for model estimation or period of analysis these need to be set up before running the code.  For example:

If the period of analysis is to be anything other than a Water Year (paStart=10, paLong=12), the user must use the `setPA` function.  For example if we want to do the analysis for the three months starting with April, the command would be `eList <- setPA(eList, paStart = 4, paLong = 3)`.

If the desired value of windowY was 10 rather than the default value of 7, then the command needed would be  `eList$INFO$windowY <- 10`, similar commands can be used for windowQ, windowS, minNumObs, minNumUncen and edgeAdjust (see help page for `modelEstimation` in the **EGRET** package).

Once all of the parameters have been set (which requires no action by the user if the defaults are to be used) the set of commands for interactive processing would be this (using the example data set Choptank_eList):

```{r eval=FALSE, echo=TRUE  } 
library(EGRET)
library(EGRETci)
eList <- Choptank_eList

#Interactive function to set up trend analysis:
caseSetUp <- trendSetUp(eList)
eBoot <- wBT(eList,caseSetUp, 
             fileName = "outputText.txt")

#Interactive save output function:
saveEGRETci(eList, eBoot, caseSetUp)


```

Two sets of outputs are created, a text file which documents the results of the WBT, and an RData file (created by the `saveEGRETci` function) that saves the relevant results as data frames for use in the `plotHistogramTrend` function (described below) or other user-specified post-processing functions. 

The output in `outputText.txt` looks like this:
```


 Choptank River    Inorganic nitrogen (nitrate and nitrite)

 Water Year

  Bootstrap process, for change from Water Year 1985 to Water Year 2010
                   data set runs from Water Year 1980 to Water Year 2011
  Bootstrap block length in days 200
  bootBreak is 39  confStop is 0.7

 WRTDS estimated concentration change is   0.444  mg/L
 WRTDS estimated flux change is          0.03593  10^6 kg/yr

Should we reject Ho that Flow Normalized Concentration Trend = 0 ? Reject Ho
 best estimate is   0.444 mg/L
  Lower and Upper 90% CIs    0.308    0.506
 also 95% CIs    0.201    0.512 
 and 50% CIs    0.397    0.472
 approximate two-sided p-value for Conc      0.05
* Note p-value should be considered to be < stated value
 Likelihood that Flow Normalized Concentration is trending up =      0.988  is trending down =     0.0125

Should we reject Ho that Flow Normalized Flux Trend = 0 ? Reject Ho
 best estimate is  0.03593 10^6 kg/year
  Lower and Upper 90% CIs   0.0211   0.0452
 also 95% CIs   0.0197   0.0456 
 and 50% CIs   0.0309   0.0390
 approximate two-sided p-value for Flux      0.05
* Note p-value should be considered to be < stated value
 Likelihood that Flow Normalized Flux is trending up = 0.988  is trending down= 0.0125

 Upward trend in concentration is highly likely 
 Upward trend in flux is highly likely
 Downward trend in concentration is highly unlikely 
 Downward trend in flux is highly unlikely
```


To run the `wBT` in batch mode the only difference from this script is that the call to `trendSetUp` should include all the needed bootstrap parameters.  So, let's assume that we want use the example data set Choptank_eList and have a PA of December, January, February and March, that we wanted to set windowY = 10 and minNumObs = 50, and all other smoothing parameters at their default values.  And we want to run the trend analysis for 1990 to 2012, use a maximum of 50 replicates, a minumum of 39 replicates, and a block length of 200.  The script would be:

```{r eval=FALSE}
library(EGRET)
library(EGRETci)
eList <- Choptank_eList
eList <- setPA(eList, paStart = 12, paLong = 4)
eList$INFO$windowY <- 10
eList$INFO$minNumObs <- 50
caseSetUp <- trendSetUp(eList, 
                        year1=1990, 
                        year2=2012, 
                        nBoot = 50, 
                        bootBreak = 39, 
                        blockLength = 200)
eBoot <- wBT(eList, caseSetUp, fileName ="outputText.txt")
saveEGRETci(eList, eBoot, caseSetUp, fileName = "output")
```

This will result in the creation of the text file called `outputText.txt`, and an RData file called `output.RData`.  The RData file will contain the objects `caseSetUp, bootOut, wordsOut, xConc, xFlux, and INFO`.  Both files will be located in the working directory.



## Histograms

The function `plotHistogramTrend` plots a histogram of all of the trend magnitudes (expressed as percentage change over the period) from the full set of replicates created by running `wBT`.  The function will plot either Flow-Normalized Concentration trends or Flow-Normalized Flux trends.  The trend magnitudes for all the replicates are stored in the eBoot list.  The first example workflow can be used to plot the two histograms: one for trends in Flow Normalized Concentration and one for Flow Normalized Flux.  

A comment about the final argument in these functions: The default is `xSeq = seq(-100,100,10)`.  This means that the bins for the histogram will run from -100 to +100 in steps of 10 (the units are percentage change).  This will almost certainly be too broad a set of bins, but the first run, with the xSeq left at its devault values (meaning that you can just leave it out of the argument list) will give you an idea of the true range that you need to consider.  Then, run it again with a much tighter range that still includes all of the replicate trends.  For example, suppose we see that the replicates seen in the first plot run tell us this: the lowest value is between -30 and -20 and the highest value is between 40 and 50, then when we give the command again we should say `xSeq = seq(-30,50,10)`.  If it turns out that all the values are much closer to zero we might use something like `xSeq = seq(-15,5,5)` so the bins are a width of 5 rather than 10.  It is also a good idea to have the full width of the histogram include zero.  For example, in the case where the values were all in the range of +20 to +70, you might use something like `xSeq = seq(0,70,10)`.  Note that if the desired output histogram needs to be smoother, then the original `wBT` run needs to set `bootBreak` to a larger value such as 100 or 200.  The larger `bootBreak` is, the smoother the histogram will be.    

Here is an example workflow where the analyst has already determined what a good xSeq argument might be.

```{r, fig.height=6, fig.width=6}
library(EGRET)
library(EGRETci)

# Example data included in package:
eList <- Choptank_eList # Example data from EGRET package
eBoot <- Choptank_eBoot
caseSetUp <- Choptank_caseSetUp

#Concentration:
plotHistogramTrend(eList, eBoot, caseSetUp,  
                   flux=FALSE, xSeq = seq(-20,60,5))

#Flux
plotHistogramTrend(eList, eBoot, caseSetUp,
                   flux=TRUE, xSeq = seq(-20,60,5))

```

Alternatively, the two plots can be shown side-by-side using a workflow like this.

Use the `par` function to set up both functions to plot side-by-side:

```{r , histExampleCombo, fig.width=7, fig.height=4}
par(mfrow=c(1,2))
plotHistogramTrend(eList, eBoot, caseSetUp, flux=FALSE,
                   printTitle=FALSE, ylim=c(0,0.07), xSeq=seq(-10,70,10))
plotHistogramTrend(eList, eBoot, caseSetUp, flux=TRUE,
                   printTitle=FALSE, ylim=c(0,0.07), xSeq=seq(-10,70,10))
# To return to figures printing in 1 row, 1 columns:
par(mfrow=c(1,1))                   
```


# Confidence Bands

## Introduction
The purpose here is to produce a graphic that looks just like those produced by `plotConcHist` or `plotFluxHist` but with the addition of confidence bands around the flow-normalized histories.  There are two versions of the workflow for confidence bands, the first just uses a single computer core which means it takes a long time to run (jobs can take an hour or more), but it is simple to run.  It uses the function `ciCalculations` and can be run interactively or in batch.  Two scripts for that are provided below.  It is best done in the Terminal or some other window, rather than in the console.  If run in the console it will make it impossible for other work in R to take place while the job is running.  The second way it can be done is using parallel computing in R.  A script for that purpose is provided.  It does require three extra packages be installed (**foreach**, **doParallel**, and **iterators**).  It is also best to run this in the Terminal, because if it is run in the console, no other R work can be done while it is running.  

## Preliminary Steps


Regardless of the approach (single core or parallel) being taken to the computations, the user needs to set up any non-default values of the parameters for model estimation and period of analysis.  For example:

If the period of analysis is to be anything other than a Water Year (paStart=10, paLong=12), the user must use the `setPA` function.  For example if we want to do the analysis for the three months starting with April, the command would be `eList <- setPA(eList, paStart = 4, paLong = 3)`.

If the desired value of windowY was 10 rather than the default value of 7, then the command needed would be  `eList$INFO$windowY <- 10`, similar commands can be used for windowQ, windowS, minNumObs, minNumUncen and EdgeAdjust (see help page for `modelEstimation`).  

It is imperative that the standard WRTDS trend analysis be run prior to running the code for the confidence bands.  If the user has made changes in smoothing parameters or modified the data in any way since the last time `modelEstimation` was run, the standard running of `modelEstimation` must be done and the output saved as `eList`.  The commands would be:

```{r, eval=FALSE, echo=TRUE}
eList <- modelEstimation(eList)
```


## Simple Workflow

This example workflow assumes that we use the default values for all of the model parameters and PA and uses the example eList provided in the package.

```{r, eval=FALSE}
library(EGRET)
library(EGRETci)

eList <- Choptank_eList

CIAnnualResults <- ciCalculations(eList)

save(eList,CIAnnualResults, file="CIAnnualResults.RData")
```

To run the `ciCalculations` function in batch mode, the only difference is that additional arguments are specified in the call to the function.  For example the call to `ciCalculations` might read as follows:

```{r eval=FALSE}
CIAnnualResults <- ciCalculations(eList, nBoot = 100, blockLength = 200, widthCI = 90)
```
As long as all required arguments (nBoot, blockLength, and widthCI) are specified there will be no interaction needed.

## Parallel Computing

Taking advantage of the **foreach** package to do parallel computing:

```{r, eval=FALSE}
library(foreach)
library(doParallel)
library(iterators)
library(EGRET)
library(EGRETci)

eList <- Choptank_eList
eList <- modelEstimation(eList) 

nBoot <- 100
blockLength <- 200
coreOut <- 1 #Number of cores to leave out of processing tasks

widthCI <- 90
ciLower <- (50-(widthCI/2))/100
ciUpper <- (50+(widthCI/2))/100
probs <- c(ciLower,ciUpper)

nCores <- detectCores() - coreOut
cl <- makeCluster(nCores)
registerDoParallel(cl)
repAnnual <- foreach(n = 1:nBoot,.packages=c('EGRETci')) %dopar% {
   annualResults <- bootAnnual(eList, blockLength)  
}
stopCluster(cl)               

# save(repAnnual, file="repAnnual.RData")

CIAnnualResults <- ciBands(eList, repAnnual, probs)
save(eList,CIAnnualResults, file="CIAnnualResults.RData")

```

## CI Result Plots

Regardless of the way that the computations are done the outputs that are needed for the graphics are contained in eList and CIAnnualResults (which in either script will end up in a file called CIAnnualResults.RData).  Once that file is loaded the graphics can be produced using the functions `plotConcHistBoot` and `plotFluxHistBoot`.  Each of these functions only require two arguments (eList and CIAnnualResults).  However, the user can specify a number of other arguments.  These other arguments are the same ones used in the `plotConcHist` and `plotFluxHist` functions in the base **EGRET** package.

The user can also view a table of the confidence interval values shown in these graphics simply by using the command `CIAnnualResults` (fluxes will be in units if kg/day).

Here is a script for producing these graphics, using the example eList and CIAnnualResults data frames included in the package.

```{r, fig.height=5, fig.width=7}
eList <- Choptank_eList

CIAnnualResults <- Choptank_CIAnnualResults

plotConcHistBoot(eList, CIAnnualResults)

plotFluxHistBoot(eList, CIAnnualResults)

```